## Evolutionary History of the Main Biological Characters of Siphonophores

### 2017-02-14 character_coding/CharacterEvolution.Rmd
"fixed graphical issues"

## Stochastic mapping plots

```{r simmap_making, echo=F, message=F, warning=F}
#Make simmap objects
make.simmap(ultratree, sex, nsim = 100) -> sex_sim
make.simmap(ultratree, palpons, nsim = 100) -> palpons_sim
make.simmap(ultratree, tentilla, nsim = 100) -> tentilla_sim
make.simmap(ultratree, nectosome, nsim = 100) -> nectosome_sim
make.simmap(ultratree, necto_pos, nsim = 100) -> necto_pos_sim
make.simmap(ultratree, necto_types, nsim = 100) -> necto_types_sim
make.simmap(ultratree, pneuma, nsim = 100, pts=T) -> pneuma_sim
make.simmap(ultratree, bracts, nsim = 100) -> bracts_sim
make.simmap(ultratree, vrange, nsim = 100) -> vrange_sim

@@

###Sex distribution:

```{r sex, echo=F, message=F, warning=F}
#Simmap SEX
plotTree(ultratree, lwd = 4)
sex_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "red")
names(colors) = c("Dioeicous", "Monoeicous")
nodelabels(pie=(describe.simmap(sex_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
densityMap(sex_sim)

###Presence of palpons:
```{r palpons, echo=F, message=F, warning=F}
#Simmap PALPONS
plotTree(ultratree, lwd = 4)
palpons_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "red")
names(colors) = c("Absent", "Present")
nodelabels(pie=(describe.simmap(palpons_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
densityMap(palpons_sim)
```

###Presence of tentilla:
```{r tentilla, echo=F, message=F, warning=F}
#Simmap TENTILLA
plotTree(ultratree, lwd = 4)
tentilla_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "red")
names(colors) = c("Absent", "Present")
nodelabels(pie=(describe.simmap(tentilla_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
densityMap(tentilla_sim)
```

###Presence of Nectophores:
```{r nectosome, echo=F, message=F, warning=F}
#Simmap NECTOSOME
plotTree(ultratree, lwd = 4)
nectosome_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "red")
names(colors) = c("Absent", "Present")
nodelabels(pie=(describe.simmap(nectosome_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
densityMap(nectosome_sim)
```

###Position of Nectosome:
```{r nectosome_position, echo=F, message=F, warning=F}
#SIMMAP NECTOSOME POSITION
plotTree(ultratree, lwd = 4)
necto_pos_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "red", "green")
names(colors) = c("Dorsal", "No nectosome", "Ventral")
nodelabels(pie=(describe.simmap(necto_pos_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
```

###Nectophore number and types:
```{r nectophore_number_types, echo=F, message=F, warning=F}
#Simmap NECTOPHORE TYPES
plotTree(ultratree, lwd = 4)
necto_types_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "red", "green", "blue")
names(colors) = c("Multiple of one type", "None", "Two of one type", "Two of two types")
nodelabels(pie=(describe.simmap(necto_types_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
```

###Presence of Pneumatophore:
```{r pneumatophore, echo=F, message=F, warning=F}
#Simmap PNEUMATOPHORE
plotTree(ultratree, lwd = 4)
pneuma_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "red")
names(colors) = c("Absent", "Present")
nodelabels(pie=(describe.simmap(pneuma_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
densityMap(pneuma_sim)
```

###Number of Bract types:
```{r bract_types, echo=F, message=F, warning=F}
#Simmap bract-types
plotTree(ultratree, lwd = 4)
bracts_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "red", "green3", "blue")
names(colors) = c("No bracts", "1", "2", "4")
nodelabels(pie=(describe.simmap(bracts_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
```

###Ocean Depth Zones:
```{r depth_zones, echo=F, message=F, warning=F}
#Simmap Vertical.Range
plotTree(ultratree, lwd = 4)
vrange_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "cyan", "blue", "green", "red")
names(colors) = c("Bathypelagic", "Surface", "Mesopelagic", "Epipelagic", "Deep benthic")
nodelabels(pie=(describe.simmap(vrange_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
```

@@

###Bathymetrics phylomorphospace
```{r depth_phylomorpho, echo=F, message=F, warning=F}
par(oma=rep(4,4))
phylomorphospace(depthmap$tree, cbind(depth_max, depth_min), xlab="Minimum depth (m)", ylab="Maximum depth (m)", colors=depthmap$cols, fsize=0.8, label = "horizontal")
```

###Bathymetrics phylomorphospace with time from root as color
```{r depth_phylochronomorpho, echo=F, message=F, warning=F}
evotime = fastBM(dpruned_tree,nsim=2)
evotime_map <- contMap(dpruned_tree,evotime[,1], plot=F)
evotime_map$cols[]<-rainbow(1001,start=0.7,end=0)
phylomorphospace(evotime_map$tree, cbind(depth_max, depth_min), xlab="Minimum depth (m)", ylab="Maximum depth (m)", colors=evotime_map$cols, fsize=0.8, label = "horizontal")
#add.color.bar(2,evotime_map$cols,title="Time from the root", lims=c(0,2000),digits=5)
```

@@

###Uncertainty inclusive traitgram for depth range
```{r uncertainty_depth, echo=F, message=F, warning=F}
contTrait <- depth_range
A<-fastAnc(dpruned_tree,contTrait,CI=TRUE)
paintree<-paintSubTree(dpruned_tree,node=length(dpruned_tree$tip)+1,"1")
trans<-as.character(floor(0:50/2))
trans[as.numeric(trans)<10]<- paste("0", trans[as.numeric(trans)<10],sep="")
# plot
for(i in 0:50){
  p<-i/length(trans)
  phenogram(dpruned_tree,c(contTrait,(1-p)*A$CI95[,1]+p*A$ace), colors=setNames(paste("#0000ff",trans[i+1],sep=""),1), add=i>0)
  phenogram(dpruned_tree,c(contTrait,(1-p)*A$CI95[,2]+p*A$ace), colors=setNames(paste("#0000ff",trans[i+1],sep=""),1), add=TRUE)
}
phenogram(dpruned_tree,c(contTrait,A$ace),add=TRUE, colors=setNames("black",1))
```

@@

###Overview of main binary traits
```{r ggtree_bin_heatmap, echo=F, message=F, warning=F}
p <- ggtree(ultratree)
p <- p + geom_tiplab(size=3)
f_cdata = sapply(cdata, as.factor) %>% as.data.frame %>% .[c(2,5:7)]
rownames(f_cdata) = rownames(cdata)
names(f_cdata)[4] = "Float"
gheatmap(p, f_cdata, offset = 0.4, width=0.5, colnames_position = 'top', font.size = 3)
```

##Phylogenetic signal in the character data
###Agnostic branch length tree generation
```{r flat_tree}
flat_tree = tree
flat_tree$edge.length = rep(mean(flat_tree$edge.length), length(flat_tree$edge.length))
```

###Phylogenetic signal in bract number
```{r phylosig_bracts}
phylosignal(bracts, ultratree)
physignal(bracts, ultratree)
```

###Phylogenetic signal in Nectosome, Palpons, Tentilla, and Pneumetophore
```{r phylosig_binaries, echo=F, message=F, warning=F}
multiPhylosignal(as.matrix(cdata[,c(2,5,6,7)]), ultratree)
```

###Phylogenetic signal in VARS bathymetrics
```{r phylosig_depth, echo=F, message=F, warning=F}
physignal(dpruned_data[,c(13,14,15,17,18)], dpruned_tree)
multiPhylosignal(dpruned_data[,c(13,14,15,17,18)], dpruned_tree)
```

###Phylogenetic PCA for bathymetrics
```{r pPCA_depth, echo=F, message=F, warning=F}
phyl.pca(dpruned_tree, dpruned_data[,c(13,14,15,17,18)], mode="corr") %>% biplot
```

###Phylogenetic ANOVA for depth
Each phylogenetic ANOVA compares variance of depth (continuous variable) within and between groups clustered by their categorical biological traits.
```{r phylANOVA_depth, echo=F, message=F, warning=F}
phylANOVA(dpruned_tree, dpruned_data$Nectosome.position, as.numeric(depth_median))
phylANOVA(dpruned_tree, dpruned_data$Sex, as.numeric(depth_median))
phylANOVA(dpruned_tree, dpruned_data$Nectosome, as.numeric(depth_median))
phylANOVA(dpruned_tree, dpruned_data$Nectophores, as.numeric(depth_median))
phylANOVA(dpruned_tree, dpruned_data$Palpons, as.numeric(depth_median))
phylANOVA(dpruned_tree, dpruned_data$Tentilla, as.numeric(depth_median))
phylANOVA(dpruned_tree, dpruned_data$Pneumatophore, as.numeric(depth_median))
```

>>>>

##Stochastic trait mapping:
```{r simmap_making, echo=F, message=F, warning=F}
#Make simmap objects
print("SIMMAP Sex Distribution")
make.simmap(ultratree, sex, nsim = 100) -> sex_sim
print("SIMMAP Palpons")
make.simmap(ultratree, palpons, nsim = 100) -> palpons_sim
print("SIMMAP Tentilla")
make.simmap(ultratree, tentilla, nsim = 100) -> tentilla_sim
print("SIMMAP Nectosome")
make.simmap(ultratree, nectosome, nsim = 100) -> nectosome_sim
print("SIMMAP Nectosome Position")
make.simmap(ultratree, necto_pos, nsim = 100) -> necto_pos_sim
print("SIMMAP Nectophore Types")
make.simmap(ultratree, necto_types, nsim = 100) -> necto_types_sim
print("SIMMAP Pneumatophore")
make.simmap(ultratree, pneuma, nsim = 100, pts=T) -> pneuma_sim
print("SIMMAP Bract Types")
make.simmap(ultratree, bracts, nsim = 100) -> bracts_sim
print("SIMMAP Habitat")
make.simmap(ultratree, vrange, nsim = 100) -> vrange_sim

@@

###SIMMAP Sex distribution:
```{r sex, echo=F, message=F,fig.height=7, fig.width=12, warning=F}
#Simmap SEX
plotTree(ultratree, lwd = 4)
sex_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "red")
names(colors) = c("Dioecious", "Monoecious")
nodelabels(pie=(describe.simmap(sex_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
densityMap(sex_sim)
```

Posterior probabilities of states (1 Dioeic)

###SIMMAP Presence of palpons:
```{r palpons, echo=F, message=F,fig.height=7, fig.width=12, warning=F}
#Simmap PALPONS
plotTree(ultratree, lwd = 4)
palpons_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "red")
names(colors) = c("Absent", "Present")
nodelabels(pie=(describe.simmap(palpons_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
densityMap(palpons_sim)
```

###SIMMAP Presence of tentilla:
```{r tentilla, echo=F, message=F,fig.height=7, fig.width=12, warning=F}
#Simmap TENTILLA
plotTree(ultratree, lwd = 4)
tentilla_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "red")
names(colors) = c("Absent", "Present")
nodelabels(pie=(describe.simmap(tentilla_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
densityMap(tentilla_sim)
```

###SIMMAP Presence of Nectophores:
```{r nectosome, echo=F, message=F,fig.height=7, fig.width=12, warning=F}
#Simmap NECTOSOME
plotTree(ultratree, lwd = 4)
nectosome_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "red")
names(colors) = c("Absent", "Present")
nodelabels(pie=(describe.simmap(nectosome_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
densityMap(nectosome_sim)
```

###SIMMAP Position of Nectosome:
```{r nectosome_position, echo=F, message=F,fig.height=7, fig.width=12, warning=F}
#SIMMAP NECTOSOME POSITION
plotTree(ultratree, lwd = 4)
necto_pos_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "red", "green")
names(colors) = c("Dorsal", "No nectosome", "Ventral")
nodelabels(pie=(describe.simmap(necto_pos_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
```

###SIMMAP Nectophore number and types:
```{r nectophore_number_types, echo=F, message=F,fig.height=7, fig.width=12, warning=F}
#Simmap NECTOPHORE TYPES
plotTree(ultratree, lwd = 4)
necto_types_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "red", "green", "blue")
names(colors) = c("Multiple of one type", "None", "Two of one type", "Two of two types")
nodelabels(pie=(describe.simmap(necto_types_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
```

###SIMMAP Presence of Pneumatophore:
```{r pneumatophore, echo=F, message=F, warning=F, fig.height=7, fig.width=12}
#Simmap PNEUMATOPHORE
plotTree(ultratree, lwd = 4)
pneuma_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "red")
names(colors) = c("Absent", "Present")
nodelabels(pie=(describe.simmap(pneuma_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
densityMap(pneuma_sim)
```

###SIMMAP Number of Bract types:
```{r bract_types, echo=F, fig.height=7, fig.width=12, message=F, warning=F}
#Simmap bract-types
plotTree(ultratree, lwd = 4)
bracts_sim %>% plotSimmap(lwd = 4, add = T) -> bract_sim_plot
#plot(bract_sim_plot[2])
colors = c("black", "red", "green3", "blue")
names(colors) = c("No bracts", "1", "2", "4")
nodelabels(pie=(describe.simmap(bracts_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
```

###SIMMAP Habitat:
```{r depth_zones, echo=F, fig.height=7, fig.width=12, message=F, warning=F}
#Simmap Vertical.Range
plotTree(ultratree, lwd = 4)
vrange_sim %>% plotSimmap(lwd = 4, add = T)
colors = c("black", "cyan", "blue", "green", "red")
names(colors) = c("Bathypelagic", "Surface", "Mesopelagic", "Epipelagic", "Deep benthic")
nodelabels(pie=(describe.simmap(vrange_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
```

@@

###Bathymetrics phylomorphospace
```{r depth_phylomorpho, echo=F, fig.height=12, fig.width=20, message=F, warning=F}
par(oma=rep(3,4))
phylomorphospace(depthmap$tree, cbind(depth_max, depth_min), xlab="Minimum depth (m)", ylab="Maximum depth (m)", colors=depthmap$cols, fsize=0.8, label = "horizontal")
```

###Bathymetrics phylomorphospace with time from root as color
```{r depth_phylochronomorpho, echo=F, fig.height=12, fig.width=20, message=F, warning=F}
evotime = fastBM(dpruned_tree,nsim=2)
evotime_map <- contMap(dpruned_tree,evotime[,1], plot=F)
evotime_map$cols[]<-rainbow(1001,start=0.7,end=0)
phylomorphospace(evotime_map$tree, cbind(depth_max, depth_min), xlab="Minimum depth (m)", ylab="Maximum depth (m)", colors=evotime_map$cols, fsize=0.8, label = "horizontal")
#add.color.bar(2,evotime_map$cols,title="Time from the root", lims=c(0,2000),digits=5)
```

@@

###Uncertainty inclusive traitgram for depth range
```{r uncertainty_depth, echo=F, fig.height=14, fig.width=14, message=F, warning=F}
contTrait <- depth_range
A<-fastAnc(dpruned_tree,contTrait,CI=TRUE)
paintree<-paintSubTree(dpruned_tree,node=length(dpruned_tree$tip)+1,"1")
trans<-as.character(floor(0:50/2))
trans[as.numeric(trans)<10]<- paste("0", trans[as.numeric(trans)<10],sep="")
# plot
for(i in 0:50){
  p<-i/length(trans)
  phenogram(dpruned_tree,c(contTrait,(1-p)*A$CI95[,1]+p*A$ace), colors=setNames(paste("#0000ff",trans[i+1],sep=""),1), add=i>0)
  phenogram(dpruned_tree,c(contTrait,(1-p)*A$CI95[,2]+p*A$ace), colors=setNames(paste("#0000ff",trans[i+1],sep=""),1), add=TRUE)
}
phenogram(dpruned_tree,c(contTrait,A$ace),add=TRUE, colors=setNames("black",1))
```

@@

###Overview grid of main binary traits
```{r ggtree_bin_heatmap, echo=F, message=F, warning=F, fig.height=10, fig.width=18}
p <- ggtree(ultratree)
p <- p + geom_tiplab(size=3)
f_cdata = sapply(cdata, as.factor) %>% as.data.frame %>% .[c(2,5:7)]
rownames(f_cdata) = rownames(cdata)
names(f_cdata)[4] = "Float"
gheatmap(p, f_cdata, offset = 0.4, width=0.5, colnames_position = 'top', font.size = 3)
```

##Phylogenetic signal in the character data
Agnostic branch length tree generation:
```{r flat_tree}
flat_tree = tree
flat_tree$edge.length = rep(mean(flat_tree$edge.length), length(flat_tree$edge.length))
```

```{r phylosig_bracts_declare, include=F, echo=F, message=F, warning=F}
physig_bracts_simp = phylosignal(bracts, ultratree)
physig_bracts_mult = physignal(bracts, ultratree)
```
###Phylogenetic signal in bract number
```{r phylosig_bracts, echo=F}
physig_bracts_simp
physig_bracts_mult
```

```{r phylosig_binaries_declare, echo=F, message=F, warning=F, include=F}
binaries_phylosig = multiPhylosignal(as.matrix(cdata[,c(2,5,6,7)]), ultratree)
```
###Phylogenetic signal in Binary Traits
```{r phylosig_binaries, echo=F, message=F, warning=F}
binaries_phylosig
```

```{r phylosig_depth_declare, message=F, warning=F, include=FALSE}
depth_physig = physignal(dpruned_data[,c(13,14,15,17,18)], dpruned_tree)
depth_multisig = multiPhylosignal(dpruned_data[,c(13,14,15,17,18)], dpruned_tree)
```
###Phylogenetic signal in VARS bathymetrics
```{r phylosig_depth, echo=F, message=F, warning=F}
depth_physig
depth_multisig
```

###Phylogenetic PCA for bathymetrics
```{r pPCA_depth, echo=F, fig.height=18, fig.width=18, message=F, warning=F}
phyl.pca(dpruned_tree, dpruned_data[,c(13,14,15,17,18)], mode="corr") %>% biplot
```

###Phylogenetic ANOVA for depth
```{r phylANOVA_depth_declare, echo=F, message=F, warning=F, include=F}
z_nectopos = phylANOVA(dpruned_tree, dpruned_data$Nectosome.position, as.numeric(depth_median))
z_sex = phylANOVA(dpruned_tree, dpruned_data$Sex, as.numeric(depth_median))
z_necto = phylANOVA(dpruned_tree, dpruned_data$Nectosome, as.numeric(depth_median))
z_nectotype = phylANOVA(dpruned_tree, dpruned_data$Nectophores, as.numeric(depth_median))
z_palp = phylANOVA(dpruned_tree, dpruned_data$Palpons, as.numeric(depth_median))
z_tentilla = phylANOVA(dpruned_tree, dpruned_data$Tentilla, as.numeric(depth_median))
z_float = phylANOVA(dpruned_tree, dpruned_data$Pneumatophore, as.numeric(depth_median))
```
Each phylogenetic ANOVA compares variance of depth (continuous variable) within and between groups clustered by their categorical biological traits.
```{r phylANOVA_depth, echo=F, message=F, warning=F}
print("Depth vs Sex Distribution")
z_sex
print("Depth vs Nectosome")
z_necto
print("Depth vs Nectophore Types and Number")
z_nectotype
print("Depth vs Nectosome Position")
z_nectopos
print("Depth vs Palpons")
z_palp
print("Depth vs Tentilla")
z_tentilla
print("Depth vs Pneumatophore")
z_float
```
